---
title: "Reproducible Research Project 1"
author: "Kintak Raymond Yu"
date: "26/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", message = FALSE, warning = FALSE)
```

## Overview 

This document summaries the exploratory data analysis of data collected from a personal activity monitoring device. 
The device records the number of steps taken in 5 minute intervals each day for an anonymous individual during October and November, 2012.

## Loading and preprocessing the data
```{r, echo = TRUE, results = "show"}
# Read in the data
library(tidyverse)
df <- read_csv("../Data/activity.csv", col_names = TRUE)
summary(df)
str(df)
```


## Number of steps taken each day
First, Let's find out how many steps are taken each day?

```{r, echo = TRUE, results = "show"}
# How many steps are taken each day? 
library(tidyverse)
dd <- df %>% group_by(date) %>% summarise(counts= sum(steps))

library(ggplot2)
Pdf <- ggplot(data = dd, aes(x = date, y = counts)) +
  geom_col() +
  labs(title="Number of steps taken each day", x = "Date", y="No. of steps")
Pdf

```

The number of steps looks rather even across the time frame. What is the mean and median of the total number of steps taken per day?

It was mentioned that there is __NA__ in the data. Let's see how they are like first.

```{r, echo=TRUE, results="show"}
library(tidyverse)
# Check NA
dfNA <- df %>% filter(is.na(steps)) %>% group_by(date) %>% summarise(`counts of NA` = n())

# list the counts
dfNA$`counts of NA`

# graph the distribution over the days
library(ggplot2)
PNA <- ggplot(data = dfNA, aes(x = date, y = `counts of NA`)) +
  geom_col()
PNA

# simple counts
nNADay <- dfNA %>% summarise(n=n())
nNATot <- sum(dfNA$`counts of NA`)
```

In the data set, there are `r nNATot` __NA__ records, evenly spread across `r nNADay` days. 

Since, 12 $\times$ 24 = `r 12*24`, so whenever there are __NA__ records in day, the data for that whole day will be missing.

Let's compute the mean in 2 ways: 

1) average the total number of steps over all the days,
2) average over only those days that does not have __NA__.

On the other hand, since the __NA__ results will not alter the median.
No such consideration is made.

```{r, echo=TRUE, results="show"}
nDay <- dd %>% summarise(n())

# mean, counting all the days
nStepsMeanAllDays   <- dd %>% summarise(sum(counts, na.rm=TRUE))/nDay
nStepsMeanAllDays

# mean, counting only the no NA days
nStepsMeanNoNADays <- dd %>% summarise(mean(counts, na.rm=TRUE))
nStepsMeanNoNADays

# median
nStepsMedianNoNADays <- dd %>% summarise(median(counts, na.rm=TRUE))
nStepsMedianNoNADays
```

Naturally, the mean values counting all the days is smaller than that only counting the no __NA__ days are different.

The mean (ignoring the days with __NA__) and the median are actually very close.

## Average daily activity pattern

Here, the average daily activity pattern is examined.
```{r, echo=TRUE}
# prep the data to get the daily average of steps
dfAvg <- df %>% group_by(interval) %>% summarise(mean = mean(steps, na.rm=TRUE))
library(ggplot2)
PAvg <- ggplot(data = dfAvg, aes(interval, mean)) +
  geom_line() + 
  labs(title="Average number of steps during each day", x = "Time interval in a day", y="Avg. no. of steps")
PAvg
# get the index of the max interval
im <- which.max(dfAvg$mean)

```

The internal that has the max value is the `r im`$^{th}$ internal, which corresponds to the interval value of `r dfAvg$interval[im]`.

```{r, echo=TRUE, results='show'}
# check the pattern of the interval to see what it means ...
head(df$interval,36)
```
Looking at the pattern in the time interval, `r dfAvg$interval[im]` should mean 8:35 am. 

## About missing values
Recall from above that there are `r nNATot` __NA__ records, evenly spread across `r nNADay` days.
In particular, when there is __NA__ in a day, the data for that whole day will be missing. 

Let's impute the missing values with the average number of steps for that 5 minute interval obtained above.
In other words, the day with __NA__ is set to be the same as the averaged day pattern.

```{r, echo=TRUE}
# create the new dff with the NA interval filled with the mean of that interval
dff <- df
dff <- dff %>% mutate (steps = ifelse(date %in% dfNA$date, dfAvg$mean, steps))
```

```{r, echo=TRUE}
# prep the data to get the total number of daily steps again
dfTot <- dff %>% group_by(date) %>% summarise(total = sum(steps))
library(ggplot2)
Pdff <- ggplot(data = dfTot, aes(date, total)) +
  geom_col() + 
  labs(title="Total number of steps in each day with NA's imputed", x = "Date", y="Total no. of steps")
Pdff
```

```{r, echo=TRUE}
# compare with the originally again
Pdf
```

```{r, echo=TRUE, results='show'}
# get the new mean and median
ddnew <- dff %>% group_by(date) %>% summarise(counts= sum(steps)) 
nStepsMeanNew <- ddnew %>% summarise(mean(counts))
nStepsMeanNew
nStepsMedianNew <- ddnew %>% summarise(median(counts))
nStepsMedianNew
```
Comparing this with the previous results, the new Mean is close to the mean when __NA__ in the data are ignored using __na.rm__. 
But it is very much different from the one computed by dividing the total steps over the total number of days (which is only `r round(nStepsMeanAllDays, 2)`).

As for the median, it is very slightly increased by the data imputing.

## Activity patterns between weekdays and weekends
```{r, echo=TRUE, results='show'}
# add a two-level factor for weekdays and weekend
library(tidyverse)
df$days <- as.factor(weekdays(df$date))
df$day2 <- fct_collapse(df$days, 
                        weekday = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'),
                        weekend = c('Saturday', 'Sunday'))

# get the average step per intervals for weekday and weekend
dfAvgWeekday <- df %>% filter(day2=='weekday') %>% 
  group_by(interval) %>% 
  summarise(mean = mean(steps, na.rm=TRUE))

dfAvgWeekend <- df %>% filter(day2=='weekend') %>% 
  group_by(interval) %>% 
  summarise(mean = mean(steps, na.rm=TRUE))

# prepare plots to compare the pattern
library(ggplot2)
library(ggpubr)
pweekday <- ggplot(data = dfAvgWeekday, aes(interval, mean)) +
  geom_line() + 
  labs(title="Average number of steps during weekday", x = "Time interval in a day", y="Avg. no. of steps")


pweekend <- ggplot(data = dfAvgWeekend, aes(interval, mean)) +
  geom_line() + 
  labs(title="Average number of steps during weekend", x = "Time interval in a day", y="Avg. no. of steps")

figure <- ggarrange(pweekday, pweekend, ncol = 1, nrow = 2)
figure

```

It is clear that during the weekend, the subject is more active during the day.
